Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-10-29 writeup/review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Local PCA shows how the effect of population structure differs along the genome''.
We appreciate the thoughtful comments by the reviewers and the editors,
and have made changes to address all of the points raised.
The changes are mostly to what is discussed;
we sympathize with the additional testing suggested by Reviewer 1,
but think that it falls outside the scope of this paper,
and would be appropriate for follow-up work that builds on our descriptive method
to create more specific, hypothesis-driven tools.

We hope you find the revised manuscript to be as improved as we do.

@@

\reply{
    We've clarified this to ``correlations in patterns of relatedness'',
    which we think isn't tautological any more,
    if LD is correlations in allelic identity, anyhow? \revref
}

\begin{point}{Figures S5 and S6:} 
    Include a panel which gives a legend/image to explain the colors being used for the bottom PCA such that the reader can know what it will look like when one of the PCAs in that row is mirroring geography.
\end{point}

\reply{
    Done.
}

\begin{point}{Senior editor comments:}
    It would be helpful to expand the discussion of variation in relationships
    along the genome - as written, the discussion is strongly focused on the
    ``classical'' Drosophila-centred linked selection literature.  Specifically,
    there has been a lot of discussion in the phylogenetics world about how to
    estimate species trees from the set of locus-specific genealogies (where the
    ``consensus'' tree can be quite wrong); and there are methods that use the
    distribution of genealogical relationships to estimate population structure
    (Konrad Lohse, Asger Hobolth etc).  It would be helpful to make these broader
    links.  
\end{point}

\reply{
    Good suggestion; we've added (hopefully helpful) additional discussion.
    {\revref} and \revreffull{1}{2}
}

@@

\reply{
    These are excellent questions for a method designed to detect inversions.
    However, we do not present such a method,
    we rather present a visualization method in which, empirically, inversions often leave a large (but not quantified) signal.
    We are not sure if it would be productive to turn this into a quantitative method, either:
    as we have now clarified in the paper \llname{ll:inversion_caveat},
    distinguishing inversions from regions of low recombination
    likely requires other sources of information.
}

\begin{point}{}
    The authors argue that linked selection is the likely cause for the patterns they observed in Drosophila dataset since the recombination rate correlates with the position of the first MDS coordinate (from line number 321). At the same time, the authors mention that the ability to identify the windows could be influenced by recombination rate in the case of linked selection and also in the presence of recombination hotspots (section 3.1 Validation). As such, I do not agree with the authors, that it is enough to demonstrate that linked selection could cause these patterns but to use this in interpretation, they should be able to distinguish this explanation from the other possible causes in simulations first and then on real data (e.g., using recombination maps for Drosophila to simulate exactly the same regions could be of advantage).
\end{point}

\reply{
    We don't share the reviewer's concern about variation in recombination rate:
    although a region of low recombination could be dominated by a single tree,
    and therefore provide a strongly divergent PCA signal,
    there is no reason this would provide a \emph{consistent} signal
    across a large fraction of a chromosome,
    as we see here.
    We've tried to make this argument more clear. {\llname{ll:not_neutral}}
    More generally, we heartily agree that we would love to have available a test
    that could prove to what degree these (or any) varying patterns of relatedness along genomes
    are due to linked selection, demographic noise, or some other force.
    However, this is one of the major unsolved question in the field,
    and is subject to substantial debate 
    (see for instance \citet{jensen2014unfounded,harris2018unfounded}
    and subsequent rejoinders).
    We've added more to the discussion on this. \revref
}

\begin{point}{}
    The authors refer to supplementary Figures sometimes as ``Figure S6'' and sometimes as ``Supplementary Figure S6''. 
    In certain journals, these two could denote different figures so I would select one way only.
\end{point}

@@

\reply{
    That's a good suggestion -- perhaps, to test for (or, visualize) outlierness,
    one could plot the deviation from the MDS centroid.
    We chose to plot each MDS coordinate separately
    so as to better see how the structure of the MDS plot lays out along the genome.
    For instance, in Figure \ref{fig:mds_allchr} on chromosome arm 3L,
    it is interesting that the green arm of the MDS plot localizes to the endpoints of the inversion,
    while the purple arm localizes to near the centromere.
    If only a measure of outlierness were plotted,
    it would be more difficult to discern patterns like this,
    and therefore develop hypotheses explaining the observed patterns.
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    The authors have put in a considerable effort to address the concerns of the reviewers. The simulation studies are now much better and the conclusions were weakened somewhat in light of them. The method definitely appears to be useful. I think that it could be an important part of the literature that moves us forward in the empirical understanding of how the genome evolves.
\end{quote}

\begin{point}{}
I only have one major comment, which is unfortunate because it could have been avoided if the simulation were implemented slightly differently. The authors didn't address the intention of my concern about the switching of PCs. To be very explicit: when you perform a PC analysis, there is some variability about the order that the ``true'' directions of the data comes out in terms of the magnitude of that PC (i.e. the order). This model uses K PCs for each small segment of genome, and each small segment is compared using a method called MDS. It is convincing that order doesn't affect comparisons of the segments *provided that all of the components are present*. However, if the magnitude of the noise is high, then it is quite plausible that the top K components in a segment of genome are not the top K for the whole genome. The simulation is useless for reassuring on this because it sets up a situation with 2 important PCs which are always included, and the claim that ``the simulation also verifies insensitivity to ordering of top PCs'' is not really true. The real data analyses are more reassuring, given the relatively similar results between K=2 and K=5, but I think the concern remains valid.

    I would like to see evidence that the windows that are found by this method are not associated with ``which'' PCs happen to be included in the top K. To do that, you could for example calculate the top $M>K$ pcs, check the correlation with the overall PCs, and then plot ``presence'' of a PC (measured in correlation in the top K) against the MDS values. It would alternatively be possible to make an argument that it will always be fine if you use a large enough number of PCs, but this is argued to have severe computational problems (an argument I don't know that I buy). It would be enough to do this for retaining *one* PC in the simulation that has 2 \emph{equally strong} PCs, but I have a suspicion that this case goes quite wrong and is anyway unrealistic. Better to keep more, e.g. 5 when there is non-trivial structure in the truth, and see that the 4th or 5th swapping with the 6th or 7th doesn't affect much. To be clear: the authors shouldn't necessarily follow this route, but they should address the underlying concern.
\end{point}

\reply{
    Apologies for not realizing the precise nature of the original concern.
    Yes, we agree that if strongly different PCs switched order 
    right at the boundary of which PCs were included,
    then that could easily show up as (spurious) structure in our MDS plots.
    However, as the reviewer says,
    spurious structure created by this problem would dissappear when using a larger number of PCs
    -- this does not happen with our empirical results.
    We have added discussion of these points. {\revref}
    Furthermore, and speculatively,
    our method of choosing a window size looks for a big enough window that the PCs within a window 
    are relatively stable,
    and so might help avoid sensitive PC switching.
}

### 2018-10-29 writeup/local_pca_paper.tex
tweaks to the responses

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
(e.g., \citet{turner2005genomic,ellegren2012genomic}).
However, a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
which in retrospect is unsurprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
The method is descriptive:
it does not aim to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome,
and give clues about its source. \llabel{ll:descriptive}
% If such variation is found, correlation of this variation with known genomic features
% can then be used to uncover its source. % \revpoint{2}{1}

@@

To see how patterns of relatedness vary in the absence of polymorphic inversions,
we performed the same analyses after removing, for each chromosome arm,
any samples carrying inversions on that arm.
In the result, shown in \Figure \ref{fig:drosophila_recomb_rate} and
Supplementary \Figure \ref{sfig:mds_allchr_noinversions},
the striking peaks associated with inversion breakpoints are gone,
and previously smaller-scale variation now dominates the MDS visualization.
For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
there is also a relatively small dip on the right end of the arm (near the centromere).
In contrast, Supplementary \Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
remaining structure is dominated by the dip near the centromere.
Without inversions, variation in patterns of relatedness shown in the MDS plots
follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
This is consistent with the hypothesis that variation
is due to selection, since the strength of linked selection increases with local gene density, 
measured in units of recombination distance.
The number of genes -- measured as the number of transcription start and end sites within each window --
was not significantly correlated with MDS coordinate ($p=0.22$).

@@

\paragraph{Genealogical noise?} \revpoint{AE}{6}
The field of phylogenetics has long had to deal with the fact that 
there can be a great number of different local phylogenies along the genome,
even between species
\citep{avise1983mitochondrial,pamilo1988relationships,hobolth2007genomic}.
Loosely, we are studying the within-species patterns that might lead to such
incomplete lineage sorting in the future.
The neutral distribution of variation in these patterns 
has been used to infer demographic history,
both between species \citep{slatkin2006concordance}
and within species \citep{beeravolu2018blockwise}.
If these distinct phylogenies are merely a result of neutral stochasticity,
there is not expected to be a correlation between local phylogeny
and other genomic features.
However, in some cases the local assortment of ancestral diversity 
and subsequent introgression between sister taxa
shows clear signs of selection, 
as for instance in the wild tomato clade \citep{pease2016phylogenomics}.

@@

Another possible contributor is recent admixture between previously separated populations,
the effects of which were not uniform across the genome due to selection. \revpoint{2}{6}
For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
\citep{harris2016genetic,juric2016strength}.
African \textit{Drosophila melanogaster} are thought to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
if selection regularly favors genes from one origin,
this could lead to substantial variation in patterns of relatedness correlated with local gene density.

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). % \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information,
or even using the variation in relatedness that we observe here,
similar to the method of \citet{beeravolu2018blockwise}. \revpoint{1}{2}

@@

Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
and how many MDS coordinates to use when describing the distance matrix between windows,
but in our exploration, changing these has not produced dramatically different results.
However, this choice could easily be important:
for instance, if the $k^\text{th}$ and $(k+1)^\text{st}$ PCs 
are sufficiently different but have similar eigenvalues,
then small amounts of noise could cause these to switch,
leading to spuriously inferred differences between windows in which one or the other
was included in the top $k$ PCs.
This does not seem to be a problem in our applications,
as changing the number of PCs did not affect the qualitative results. \revpoint{2}{1}
These choices are all part of more general techniques in dimension reduction and high-dimensional data visualization;
we encourage the user to experiment.

>>>>

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
(e.g., \citet{turner2005genomic,ellegren2012genomic}).
However, a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
which in retrospect is unsurprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
The method is descriptive:
it does not aim to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome,
and give clues about the source of this variation. \llabel{ll:descriptive}
% If such variation is found, correlation of this variation with known genomic features
% can then be used to uncover its source. % \revpoint{2}{1}

@@

To see how patterns of relatedness vary in the absence of polymorphic inversions,
we performed the same analyses after removing, for each chromosome arm,
any samples carrying inversions on that arm.
In the result, shown in \Figure \ref{fig:drosophila_recomb_rate} and
Supplementary \Figure \ref{sfig:mds_allchr_noinversions},
the striking peaks associated with inversion breakpoints are gone,
and previously smaller-scale variation now dominates the MDS visualization.
For instance, the majority of the variation along 3L in \Figure \ref{fig:mds_allchr}
is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
there is also a relatively small dip on the right end of the arm (near the centromere).
In contrast, Figure \ref{fig:drosophila_recomb_rate} 
and Supplementary \Figure \ref{sfig:mds_allchr_noinversions} 
show that after removing polymorphic inversions,
remaining structure is dominated by the dip near the centromere.
Without inversions, variation in patterns of relatedness shown in the MDS plots
follows similar patterns to that previously seen in \textit{D.~melanogaster} recombination rate and diversity \citep{langley2012genomic,mackay2012drosophila}.
Indeed, correlations between the recombination rate in each window and the position on the first MDS coordinate are highly significant
(Spearman's $\rho=0.54$, $p<2 \times 10^{-16}$; \Figures \ref{fig:drosophila_recomb_rate} and \ref{sfig:drosophila_gene_density_correlations}).
This is consistent with the hypothesis that variation
is due to selection, since the strength of linked selection increases with local gene density, 
measured in units of recombination distance.
The number of genes -- measured as the number of transcription start and end sites within each window --
was not significantly correlated with MDS coordinate ($p=0.22$).

@@

\paragraph{Genealogical noise?} \revpoint{AE}{6}
The field of phylogenetics has long had to deal with the fact that 
there can be a great number of different local phylogenies along the genome,
even between species
\citep{avise1983mitochondrial,pamilo1988relationships,hobolth2007genomic}.
The within-species patterns we observe might contribute to such
incomplete lineage sorting among future descendant species of a given population.
The neutral distribution of variation in these patterns 
has been used to infer demographic history,
both between species \citep{slatkin2006concordance}
and within species \citep{beeravolu2018blockwise}.
If these distinct phylogenies are merely a result of neutral stochasticity,
there is not expected to be a correlation between local phylogeny
and other genomic features.
However, in some cases the local assortment of ancestral diversity 
and subsequent introgression between sister taxa
shows clear signs of selection, 
as for instance in the wild tomato clade \citep{pease2016phylogenomics}.

@@

Another possible contributor is recent admixture between previously separated populations,
the effects of which were not uniform across the genome due to selection. % \revpoint{2}{6}
For instance, it has been hypothesized that large-scale variation in amount of introgressed Neanderthal DNA along the genome
is due to selection against Neanderthal genes, leading to greater introgression in regions of lower gene density
\citep{harris2016genetic,juric2016strength}.
African \textit{Drosophila melanogaster} are thought to have a substantial amount of recently introgressed genome from ``cosmopolitan'' sources;
if selection regularly favors genes from one origin,
this could lead to substantial variation in patterns of relatedness correlated with local gene density.

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). % \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid 
in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations 
will affect these predictions.
Developing a method to distinguish these forms of selection from each other
and from the effects of demography is a major challenge to the field.
It may be possible to make progress using statistics that make stronger use of spatial information,
such as the variation in relatedness that we observe here,
similar to the method of \citet{beeravolu2018blockwise}. \revpoint{1}{2}

@@

Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
and how many MDS coordinates to use when describing the distance matrix between windows,
but in our exploration, changing these has not produced dramatically different results.
However, this choice could in some situations be important:
for instance, if the $k^\text{th}$ and $(k+1)^\text{st}$ PCs 
are sufficiently different but have similar eigenvalues,
then small amounts of noise could cause these to switch,
leading to spuriously inferred differences between windows in which one or the other
was included in the top $k$ PCs.
This does not seem to be a problem in our applications,
as changing the number of PCs did not affect the qualitative results. \revpoint{2}{1}
These choices are all part of more general techniques in dimension reduction and high-dimensional data visualization;
we encourage the user to experiment.
